---
title: "How Does a Bike-Share Navigate Speedy Success ?"
author: "Mahmoud Essam"
date: "`r Sys.Date()`"
output:
  html_document: default
  word_document: default
  pdf_document: default
---

## Introduction :

*Welcome to the Cyclistic bike-share analysis! in this scenario will discuess casual riders and annual members use Cyclistic bikes differently till we can convert casual riders into annual members.*

## Scenario

*The director of marketing believes the company's future success depends on maximizing the number of annual memberships Therefore, your team wants to understand how casual riders and annual members use Cyclistic bikes differently. From these insights, your team will design a new marketing strategy to convert casual riders into annual members*

## Installing packages

*first of all we will now import the important packages we need for now .*

**\# Install required packages**

**\# tidyverse for data import and wrangling**

**\# lubridate for date functions**

**\# ggplot for visualization**

## Importing Packages

```{r}
library(tidyverse)
library(lubridate)
library(janitor)
library(skimr)
library(hydroTSM)
```

## **Let's involve into the project**

*► download the data we need from [this link](https://divvy-tripdata.s3.amazonaws.com/index.html) as it's available by Motivate International Inc. under this [license](https://ride.divvybikes.com/data-license-agreement)*

► after this step now it's time to put all data into R

```{r}

d1 <- read.csv("C:/Users/BLU-RAY/Desktop/Divvy-TripData/Work_Data_Sheets_CSV/Trip_Travles_1_To_12_From_2020_To_2022/Trips_of _last_12_months/202109-divvy-tripdata.csv")
d2 <-  read.csv("C:/Users/BLU-RAY/Desktop/Divvy-TripData/Work_Data_Sheets_CSV/Trip_Travles_1_To_12_From_2020_To_2022/Trips_of _last_12_months/202110-divvy-tripdata.csv")
d3 <- read.csv("C:/Users/BLU-RAY/Desktop/Divvy-TripData/Work_Data_Sheets_CSV/Trip_Travles_1_To_12_From_2020_To_2022/Trips_of _last_12_months/202111-divvy-tripdata.csv")
d4 <- read.csv("C:/Users/BLU-RAY/Desktop/Divvy-TripData/Work_Data_Sheets_CSV/Trip_Travles_1_To_12_From_2020_To_2022/Trips_of _last_12_months/202112-divvy-tripdata.csv")
d5 <-  read.csv("C:/Users/BLU-RAY/Desktop/Divvy-TripData/Work_Data_Sheets_CSV/Trip_Travles_1_To_12_From_2020_To_2022/Trips_of _last_12_months/202201-divvy-tripdata.csv")
d6 <- read.csv("C:/Users/BLU-RAY/Desktop/Divvy-TripData/Work_Data_Sheets_CSV/Trip_Travles_1_To_12_From_2020_To_2022/Trips_of _last_12_months/202202-divvy-tripdata.csv")
d7 <-  read.csv("C:/Users/BLU-RAY/Desktop/Divvy-TripData/Work_Data_Sheets_CSV/Trip_Travles_1_To_12_From_2020_To_2022/Trips_of _last_12_months/202203-divvy-tripdata.csv")
d8 <- read.csv("C:/Users/BLU-RAY/Desktop/Divvy-TripData/Work_Data_Sheets_CSV/Trip_Travles_1_To_12_From_2020_To_2022/Trips_of _last_12_months/202204-divvy-tripdata.csv")
d9 <- read.csv("C:/Users/BLU-RAY/Desktop/Divvy-TripData/Work_Data_Sheets_CSV/Trip_Travles_1_To_12_From_2020_To_2022/Trips_of _last_12_months/202205-divvy-tripdata.csv")
d10 <- read.csv("C:/Users/BLU-RAY/Desktop/Divvy-TripData/Work_Data_Sheets_CSV/Trip_Travles_1_To_12_From_2020_To_2022/Trips_of _last_12_months/202206-divvy-tripdata.csv")
d11 <-  read.csv("C:/Users/BLU-RAY/Desktop/Divvy-TripData/Work_Data_Sheets_CSV/Trip_Travles_1_To_12_From_2020_To_2022/Trips_of _last_12_months/202207-divvy-tripdata.csv")
d12 <- read.csv("C:/Users/BLU-RAY/Desktop/Divvy-TripData/Work_Data_Sheets_CSV/Trip_Travles_1_To_12_From_2020_To_2022/Trips_of _last_12_months/202208-divvy-tripdata.csv")
```

► **Now we need to bind all this data into one**

```{r}
Ddata <- rbind(d1,d2,d3,d4,d5,d6,d7,d8,d9,d10,d11,d12)
```

► **make a great view for our current data :**

```{r}
summary(Ddata)
glimpse(Ddata)
skim_without_charts(Ddata)
```

► **Cleaning our data :**

```{r}
old_NOR <- nrow(Ddata) #to see the number of old data
New_Data <- distinct(Ddata) # to return data without any duplication
New_NOR <- nrow(New_Data) #to see the number after get duplicated
if (old_NOR == New_NOR) {
  print("This data doesn't have any duplicate numbers")
}else{
  print("This data have duplicated number = " + old_NOR-New_NOR)
}
```

► **Now we need to return all data we need to use :**

*Now will start our time with analyze :*

```{r}
unique(Ddata$rideable_type) # to check the unique kinds of rideables 
unique(Ddata$member_casual) # to check the unique kinds of riders
Ddata <- New_Data %>% select(2, 3, 4, 13) # the current data required 
Ddata$date <- as.Date(Ddata$started_at)#converting it into Date type
Ddata <- Ddata %>% mutate(
      start_year = year(started_at),
      start_month = month(started_at),
      start_day = day(started_at))
Ddata <- Ddata %>%arrange(date) #sorting the data by date
Ddata <- Ddata %>% mutate(season = time2season(date,out.fmt = "seasons")) # Convert dates to seasons
Ddata <- Ddata %>% mutate(Months = month.name[start_month]) #create month named
Ddata$day_of_week <- format(as.Date(Ddata$date), "%A") #adding day column
Ddata$day_of_week <- ordered(Ddata$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday")) # ordering day of the week
```

*But what about if our end date lower than start one ? let's check it !*

```{r}
Ddata %>% filter(ended_at < started_at) %>% count() #here will know about the number of negative date 
Ddata <- Ddata %>% filter(ended_at>started_at) #assigned the new data without any negative inputs
str(Ddata)
```

*Now back to our analyze by getting the ride length*

```{r}
Ddata <- Ddata %>% mutate(Ride_Length = as.numeric(difftime(Ddata$ended_at,Ddata$started_at)))  # to return the ride length of the bikers in seconds
Ddata <- Ddata %>% mutate(Ride_Length_in_mins = as.numeric(difftime(Ddata$ended_at,Ddata$started_at,units = "mins"))) # to return the ride length of the bikers in minutes
colSums(is.na(Ddata)) #to check the na data
```

*Give a look to our current data to make sense of it :*

```{r}
summary(Ddata$Ride_Length) #to know what's the maximum and minimum and other details of our ride
aggregate(Ddata$Ride_Length ~ Ddata$member_casual, FUN = max)# Comparing members and 
aggregate(Ddata$Ride_Length ~ Ddata$member_casual, FUN = min)# Comparing members and casual users min
aggregate(Ddata$Ride_Length ~ Ddata$member_casual, FUN = median)# Comparing members and casual users median
aggregate(Ddata$Ride_Length ~ Ddata$member_casual, FUN = mean) # Comparing members and casual users mean

```

*Now we got a great summary of our current data , now let's pivot it ! :*

```{r echo=TRUE}
#Step 1 let's know number of rides per day of member and casual 
number_of_rides_per_day <- Ddata%>%group_by(member_casual,Ddata$day_of_week)%>%summarise(number_of_rides_general=n()) 
number_of_rides_per_day
#Step 2 let's know number of rides per month of member and casual 
number_of_rides_per_month <- Ddata%>%group_by(member_casual,Ddata$Months)%>%summarise(number_of_rides_general=n()) 
number_of_rides_per_month
#step 3 let's know number of rides with kind of the ridable 
number_of_rides_per_ridable <- Ddata %>%group_by(member_casual,Ddata$rideable_type)%>%summarise(number_of_rides_general=n())
number_of_rides_per_ridable
#step 4 let's know number of rides with kind of the ridable 
number_of_rider_per_season <- Ddata%>%group_by(member_casual,Ddata$season)%>%summarise(number_of_rides_general=n())
number_of_rider_per_season
#step 5 Number of riders in last 12 months
number_of_riders_per_in_last_12_months <- Ddata%>%group_by(member_casual)%>%summarise(number_of_rides_general=n())
```

*after knowing current requirements of riders to know what's the things affect on them, let's test their time in riding :*

```{r}
#Calculate ride length per member and casual in day
Avg_Time_per_day <- aggregate(Ddata$Ride_Length_in_mins ~Ddata$member_casual + Ddata$day_of_week,FUN = mean)
Avg_Time_per_day
#Calcualte ride length per member and casual in month 
Avg_Time_per_month <- aggregate(Ddata$Ride_Length_in_mins~Ddata$member_casual + Ddata$Months,FUN = mean)
Avg_Time_per_month
#Calculate ride length per member and casual in type of ridable 
Avg_Time_per_Ride <- aggregate(Ddata$Ride_Length_in_mins~Ddata$member_casual + Ddata$rideable_type,FUN = mean)
Avg_Time_per_Ride
#Calculate ride length per member and casual in season
Avg_Time_per_season <- aggregate(Ddata$Ride_Length_in_mins~Ddata$member_casual+Ddata$season,FUN = mean)
Avg_Time_per_season
```

*in our last destination now it's time for some little visualize so we are in need to export this data now for tableau for some viz with excel !*

```{r}
#Exporting our data out 
write.csv(Ddata,"C:/Users/BLU-RAY/Desktop/Divvy-TripData/Result Data/Total_Data.csv")
write.csv(number_of_rides_per_day,"C:/Users/BLU-RAY/Desktop/Divvy-TripData/Result Data/number_of_rides_per_day.csv")
write.csv(number_of_rides_per_month,"C:/Users/BLU-RAY/Desktop/Divvy-TripData/Result Data/number_of_rides_per_month.csv")
write.csv(number_of_rides_per_ridable,"C:/Users/BLU-RAY/Desktop/Divvy-TripData/Result Data/number_of_rides_per_ridable.csv")
write.csv(number_or_rider_per_season,"C:/Users/BLU-RAY/Desktop/Divvy-TripData/Result Data/number_of_rides_per_season.csv")
write.csv(number_of_riders_per_in_last_12_months,"C:/Users/BLU-RAY/Desktop/Divvy-TripData/Result Data/number_of_riders_per_in_last_12_months.csv")
write.csv(Avg_Time_per_day,"C:/Users/BLU-RAY/Desktop/Divvy-TripData/Result Data/Avg_time_per_day.csv")
write.csv(Avg_Time_per_month,"C:/Users/BLU-RAY/Desktop/Divvy-TripData/Result Data/Avg_time_per_month.csv")
write.csv(Avg_Time_per_Ride,"C:/Users/BLU-RAY/Desktop/Divvy-TripData/Result Data/Avg_Time_per_Ride.csv")
write.csv(Avg_Time_per_season,"C:/Users/BLU-RAY/Desktop/Divvy-TripData/Result Data/Avg_Time_per_Season.csv")
```
